Demo: Counterpoint Analysis and Synthesis

### 2020-04-21 doc/icfp20/response.txt

We would like to thank the reviewers for their thorough reading and
valuable feedback.  The reviewers are right to point out our limited
use of dependent types; ...

Below is our answer to other questions.

### To Reviewer A

### To Reviewer B

* The best way to understand Line 315 is to look at the example in
  Figure 4.  In bar 5, the counterpoint line moves downwards from
  C to E (unison).  Then, in the next bar, it moves upwards from E to
  F.  We call a motion "step-wise" when its endpoints are adjacent
  notes, such as E-F.

### To Reviewer C

* The word "unspecified" in Line 87 is not a typo.  The unit of time is
  specified when the music is played at a particular tempo.

### 2020-04-22 doc/icfp20/response.txt

We would like to thank the reviewers for their thorough reading and
valuable feedback. We agree with the reviewers that the use of
dependent types in the paper is not sufficient to demonstrate their
true value in the context of music theory. There is additional use of
dependent types in the code that did not make it into the paper due to
space limitations and the difficulty of fitting it into the main
narrative, but likely this would not have been enough to change
anyone's mind. We have much more planned that should demononstrate how
compelling dependent types are in this context, but this is still
future work.

All reviewers raise other interesting points that we intend to
address, and we thank them for pointing these out. Reviewers B and C
raise doubts about whether the extrinsic approach in this paper is
really superior to the intrinsic approach of our previous
work. Although our experience so far has indicated the extrinsic
approach seems both more flexible and more user-friendly, a more
detailed comparison is warranted. We were unaware that Agda's
reflection mechanism allows custom error messages, so thanks to
Reviewer B for mentioning this and we will try it out. It is enticing
to take advantage of the built-in type checker as much as possible,
but it can also be constraining if one wants to for example explore
multiple rule systems at once. The two approaches are not so different
as indices can be represented as predicates attached to a sigma
(record) type, and with an appropriate equivalence can be converted
back and forth, which we also intend to explore. As Reviewer C notes
we need to be more clear about what advantages there are to running
the check at compile time rather than run-time in Haskell.

Reviewers B and C also mention constraint-based search in which the type
system may be used to partially constrain the search, and this is also
planned future work.

Reviewer A suggests implementing previoudTriads as a relation instead
of a function, and the tradeoff between these approaches, and indeed
many other aspects of practical programming using dependent types, are
areas we hope to explore in detail going forward.

[Note to Youyou: We could just leave out specific answers to questions
unless you think the reviewers are really interested in this.]

### 2020-04-23 doc/icfp20/response.txt

We would like to thank the reviewers for their thorough reading and
valuable feedback. We agree with the reviewers that the use of
dependent types in the paper is not sufficient to demonstrate their
true value in the context of music theory. There is additional use of
dependent types in the code that did not make it into the paper due to
space limitations and the difficulty of fitting it into the main
narrative, but likely this would not have been enough to change
anyone's mind. We have many more plans that should demonstrate how
compelling dependent types are in this context, but this is still
future work.

All reviewers raise other interesting points that we intend to
address, and we thank them for pointing these out.  Below we would
like to respond to the main points.

### Extrinsic vs. intrinsic approaches (Reviewers B and C)

The reviewers raise doubts about whether the extrinsic approach
in this paper is really superior to the intrinsic approach of previous
work. Although our experience so far has indicated the extrinsic
approach seems both more flexible and more user-friendly, a more
detailed comparison is warranted. We were unaware that Agda's
reflection mechanism allows custom error messages, so thanks to
Reviewer B for mentioning this and we will try it out. It is enticing
to take advantage of the built-in type checker as much as possible,
but it can also be constraining if one wants to for example explore
multiple rule systems at once. The two approaches are not so different
as indices can be represented as predicates attached to a $\Sigma$
(record) type, and with an appropriate equivalence can be converted
back and forth, which we also intend to explore. As Reviewer C notes
we need to be more clear about what advantages there are to running
the check at compile time rather than run-time in Haskell.

### Constraint-based search (Reviewers B and C)

The reviewers also mention constraint-based search in which the type
system may be used to partially constrain the search.  This is also
planned future work.

### Triads as a relation (Reviewer A)

The reviewer suggests implementing previousTriads as a relation instead
of a function. The tradeoff between these approaches, and indeed
many other aspects of practical programming using dependent types, are
areas we hope to explore in detail going forward.

@@

[Note to Youyou: We could just leave out specific answers to questions
unless you think the reviewers are really interested in this.]

[Note to John: I agree that these questions are not very important,
but I would prefer not to ignore them.]

Lastly, let us answer some minor questions.

@@

### To Reviewer B

* The reviewer asks how to interpret the sentence in Line 315.  The
  best way to understand it is to look at the example in Figure 4.
  In bar 5, the counterpoint line moves downwards from C to E (unison).
  Then, in the next bar, it moves upwards from E to F. We call a motion
  "step-wise" when its endpoints are adjacent notes, such as E-F.

* The reviewers wonders in what way harmony is dual to counterpoint.
  As Cong and Leo [FARM '19] suggest, we may view harmony as a
  horizontal view of a musical score, and counterpoint as a vertical view.

### 2020-04-23 doc/icfp20/response.txt

We would like to thank the reviewers for their thorough reading and
valuable feedback. We agree with the reviewers that the use of
dependent types in the paper is not sufficient to demonstrate their
true value in the context of music theory. There is additional use of
dependent types in the code that did not make it into the paper due to
space limitations and the difficulty of fitting it into the main
narrative, but likely this would not have been enough to change
anyone's mind. We have many more plans that should demonstrate how
compelling dependent types are in this context, but this is still
future work.

All reviewers raise other interesting points that we intend to
address, and we thank them for pointing these out.  Below we would
like to respond to the main points.

### Extrinsic vs. intrinsic approaches (Reviewers B and C)

The reviewers raise doubts about whether the extrinsic approach
in this paper is really superior to the intrinsic approach of previous
work. Although our experience so far has indicated the extrinsic
approach seems both more flexible and more user-friendly, a more
detailed comparison is warranted. We were unaware that Agda's
reflection mechanism allows custom error messages, so thanks to
Reviewer B for mentioning this and we will try it out. It is enticing
to take advantage of the built-in type checker as much as possible,
but it can also be constraining if one wants to for example explore
multiple rule systems at once. The two approaches are not so different
as indices can be represented as predicates attached to a sigma
(record) type, and with an appropriate equivalence can be converted
back and forth, which we also intend to explore. As Reviewer C notes
we need to be more clear about what advantages there are to running
the check at compile time rather than run-time in Haskell.

@@

Lastly, let us answer some minor questions.

### To Reviewer B

* The reviewer asks how to interpret the sentence in Line 315.  The
  best way to understand it is to look at the example in Figure 4.
  In bar 5, the counterpoint line moves downwards from C to E (unison).
  Then, in the next bar, it moves upwards from E to F. We call a motion
  "step-wise" when its endpoints are adjacent notes, such as E-F.

* The reviewer wonders in what way harmony is dual to counterpoint.
  As Cong and Leo [FARM '19] suggest, we may view harmony as a
  horizontal view of a musical score, and counterpoint as a vertical view.

### 2022-08-10 doc/farm22/abstract.tex
work on paper

\setcopyright{none}
\acmPrice{}
\acmDOI{}
\acmYear{}
\copyrightyear{}
\acmISBN{}
\acmConference[FARM '22]{}
\acmBooktitle{}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
We present a Haskell library to help analyze
and synthesize musical counterpoint. The tool allows expression of
generic constraints in a higher-level musical language which are
translated to a lower level for use both to find rule violations in
existing music and to generate (using an SMT solver) new
music satisfying the constraints. The tool is intended for use by
musicians who need only have a basic knowledge of Haskell.
\end{abstract}

@@

%% Keywords
%% comma separated list
\keywords{Counterpoint, Haskell, SMT}  %% \keywords are mandatory in final camera-ready submission

@@

Another approach then is to write a special-purpose ``type checker''
which can be run on previously-created music and which can generate
clear and precise error messages which are musically meaningful. The
checker is ideally easily customizable in terms of what constraints
one would like to impose. It turns out these constraints can be
expressed (at a low level) in the quantifier-free logic of linear
arithmetic and uninterpreted functions (QF-UFLIA \cite{Barrett2010}),
which allows one to not only analyze existing counterpoint, but also
synthesize counterpoint satisfying the constraints using an SMT
solver.

This tool, still a work in progress, has been implemented in Haskell
and is available at \cite{HaskellCounterpoint}. Haskell was chosen as
a high-quality and feature-rich SMT library called SBV \citep{SBV} is
available. The tool makes heavy use of functional programming, but
dependent types are sometimes wished for and the goal is to eventually
port the work to the Agda Music Tools \cite{MusicTools} library.

The following sections give an overview of current
functionality. Future plans include integration with Liquid Haskell
\citep{LiquidHaskell} (for analysis) and Synquid \citep{Synquid} (for
synthesis). In addition to handling higher species we would like to
also incorporate the rules and conventions of galant schemata
\citep{Gjerdingen2007}, which informed much of the music composed in
the era of Haydn and Mozart. Ideally one could use the tool to compose
convincing music in the galant style.

Concurrently Cong \citep{Cong2022a} is exploring another approach
using a different representation of constraints within a type system.

@@

There are boundary rules that the top and bottom voices must form
octaves at the beginning and end, and motion rules which are designed
to ensure independence of voices. In particular parallel and similar
(both voices moving in the same direction) motion into fifths and
octaves is prohibited. Here we can see that Beethoven has in fact made
two errors in the top two voices: similar motion into a fifth in bar 3
and then into an octave between bars 3 and 4. Haydn fixes both errors
by changing the F in the second voice to an A.

@@

At a high level music is represented as a list of voices, which in
turn are a list of bars, which are in turn a list of beats. Here
dependent types would be useful to ensure for example that every voice
has the same number of bars, but a run-time check can also be used to
enforce this. Each of these elements is parameterized, allowing a
number of pitch-like elements to be used to construct music. These
could be algebraic data types representing name (like C\#) and
octave, or a numeric representation of pitch. Either datum can then be
embedded into a wrapper, such as \texttt{MPitch} which can either be a
fixed pitch or a named hole to be solved for.

@@

@@

Thanks to Youyou Cong for many enlightening discussions about type
theory and music over the past several years, and for suggesting to
look at Beethoven's exercises as a source of examples. Thanks also to
Emina Torlak and Stephen Rumph for their generosity in allowing me to
audit their classes at the University of Washington. CSE 507 was
invaluable for learning the fundamentals of SMT 
solvers, and MUHST 211 and 430 introduced me to galant schemata.

>>>>

\setcopyright{none}
\acmPrice{}
\acmDOI{}
\acmYear{}
\copyrightyear{}
\acmISBN{}
\acmConference[FARM '22]{Proceedings of the 10th ACM SIGPLAN International Workshop on Functional Art, Music, Modeling, and Design}{September 15, 2022}{Ljubljana, Slovenia}
\acmBooktitle{}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
We present an Agda library to help analyze
and synthesize musical counterpoint. The tool allows expression of
generic constraints in a higher-level musical language which are
translated to a lower level for use both to find rule violations in
existing music and to generate (using an SMT solver) new
music satisfying the constraints. The tool is intended for use by
musicians who need only have a basic knowledge of Agda.
\end{abstract}

@@

%% Keywords
%% comma separated list
\keywords{Counterpoint, Agda, SMT}  %% \keywords are mandatory in final camera-ready submission

@@

Another approach then is to write a special-purpose ``type checker''
which can be run on previously-created music and which can generate
clear and precise error messages which are musically meaningful. The
checker is ideally easily customizable in terms of what constraints
one would like to impose. It turns out these constraints can be
expressed (at a low level) in the quantifier-free logic of linear
arithmetic and uninterpreted functions (QF-UFLIA \cite{Barrett2010}),
which allows one to not only analyze existing counterpoint, but also
synthesize counterpoint satisfying the constraints using an SMT
(Satisfiablity Modulo Theories) solver.

The initial prototype \citep{HaskellCounterpoint} of this tool was
written in Haskell, chosen due to the availability of the
high-quality and feature-rich SMT library SBV \citep{SBV}. It was then
ported to the Agda Music Tools library \cite{MusicTools}, using Agda's
Haskell foreign function interface (FFI) to call SBV. There is also an
Agda interface \citep{Schmitty} to SMT and we might convert to using
that at some point in the future.

The following sections give an overview of current
functionality.

@@

There are boundary rules that the top and bottom voices must form
octaves at the beginning and end, and motion rules which are designed
to ensure independence of voices. In particular parallel and similar
(both voices moving in the same direction) motion into fifths and
octaves is prohibited. Here we can see that Beethoven has in fact made
two errors in the top two voices: similar motion into a fifth in bar 3
and then into an octave between bars 3 and 4. Haydn fixes both errors
by changing the F in the second voice (marked with the red X) to an A.

@@

At a high level music is represented as a list of voices, which in
turn are a list of bars, which are in turn a list of beats. Here
dependent types would be useful to ensure for example that every voice
has the same number of bars, but a run-time check can also be used to
enforce this. Each of these elements is parameterized, allowing a
number of pitch-like elements to be used to construct music. These
could be algebraic data types representing name (like G\#) and
octave, or a numeric representation of pitch. Either datum can then be
embedded into a wrapper, such as \texttt{MPitch} which can either be a
fixed pitch or a named hole to be solved for.

@@

\section{Future Plans}

Future plans include possible integration with Liquid Haskell
\citep{LiquidHaskell} (for analysis) and Synquid \citep{Synquid} (for
synthesis). In addition to handling higher species we would like to
also incorporate the rules and conventions of galant schemata
\citep{Gjerdingen2007}, which informed much of the music composed in
the era of Haydn and Mozart. Ideally one could use the tool to compose
convincing music in the galant style.

Concurrently Cong \citep{Cong2022a,Cong2022c} is exploring another approach
using a different representation of constraints within a type system.
Recent work by Tanaka \citep{Tanaka2022} uses integer programming to
express constraints, and we plan to compare our method with this.

@@

Thanks to Youyou Cong for many enlightening discussions about type
theory and music over the past several years, and for suggesting to
look at Beethoven's exercises as a source of examples. Thanks also to
Emina Torlak and Stephen Rumph for their generosity in allowing me to
audit their classes at the University of Washington. CSE 507 was
invaluable for learning the fundamentals of SMT 
solvers, and MUHST 211 and 430 introduced me to galant schemata.
Finally, thanks the anonymous referees for many helpful comments and
suggestions.
